---
title: "NZ generation share trends"
subtitle: "Electricity Authority (EA) data"
author: "Ben Anderson (dataknut@icloud.com)"
date: 'Last run at: `r Sys.time()`'
output: 
  bookdown::html_document2:
    fig_caption: yes
    toc: TRUE
    toc_depth: 4
    toc_float: TRUE
    code_folding: hide
bibliography: '`r path.expand("~/bibliography.bib")`'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) # but folded by yaml setting

library(dkUtils) # https://github.com/dataknut/dkUtils

myLibs <- c("data.table", # fast data munching
            "lubridate", # time stuff
            "flextable", # pretty tables
            "ggplot2", # plots obvs
            "here", # where are we?
            "hms", # more time stuff
            "plotly"
            )
dkUtils::loadLibraries(myLibs) # load the libs (installs if not present)

# local functions ----

# project/repo parameters ----
source(here::here("env.R"))

rmdParams <- list()

rmdParams$plotPath <- here::here("rmd", "plots", "nz")

if(!dir.exists(rmdParams$plotPath)){
            dir.create(rmdParams$plotPath)
          }
```

# About

Code copyright (c) `r lubridate::year(now())` the [author](https://www.linkedin.com/in/dataknut/)

License: 
 
 * [CC-BY](https://creativecommons.org/licenses/by/4.0/legalcode)

To cite:

 * Anderson, B, (`r lubridate::year(now())`) New Zealand generation share trends: Electricity Authority data. https://dataknut.github.io/gridCarbon/nzGenMixTrends.html

Code:

 * https://github.com/dataknut/gridCarbon
 
# Introduction

Inspired by https://twitter.com/DrSimEvans/status/1508409309775994892

# Data

This analysis uses the [New Zealand electricity Authority Generation](https://www.emi.ea.govt.nz/Wholesale/Datasets/Generation/Generation_MD/) data. 

The data contains average (mean) MW generation per half hour by source including. Where necessary we convert this to MWh per half hour (MWh = mean half-hourly MW/2)

As far as we can work out this data does _not_ include distributed (i.e. non-grid connected) generation such as small scale wind, solar, hydro, biomass etc which is connected to the LV network. This means the EA data is likely to _underestimate_ total generation and potentially underestimate the proportion of total generation that is renewable. It is possible that this could be fixed using embedded wind & solar generation data from the [metered embedded generation data](https://www.emi.ea.govt.nz/Wholesale/Datasets/Metered_data/Embedded_generation) although this may still not include household level (micro)generation.

The data also includes a half-hourly carbon intensity value in g CO2/kWh sourced from https://www.carbonintensity.org.uk/.

```{r setOlderThan}
rmdParams$olderThan <- 7
```

If the data we have previously downloaded is more than `r rmdParams$olderThan`` days old, re-download.

```{r loadData}

rmdParams$dataPath <- "~/Dropbox/data/NZ_ElecAuth/processed/"

# load all the files we have using rbindlist
# probably ought to do some filtering here
filesToLoad <- list.files(rmdParams$dataPath, pattern = ".csv", 
                          full.names = TRUE)

orig_DT <- rbindlist(lapply(filesToLoad, fread))
setnames(orig_DT, "rDateTime", "dv_dateTime")

message("Original data range from: ", min(orig_DT$dv_dateTime, na.rm = TRUE))
message("...to: ", max(orig_DT$dv_dateTime , na.rm = TRUE))

# make wide for ease of aggregation ----
nzGenMix_wide_dt <- dcast(orig_DT[!is.na(dv_dateTime)], dv_dateTime ~ Fuel_Code, value.var = "kWh", fun.aggregate = sum)

# add derived variables used later ----
addDerivedVars <- function(dt){
  dt[, dv_year := lubridate::year(dv_dateTime)]
dt[, dv_date := lubridate::as_date(dv_dateTime)]
dt[, dv_month := lubridate::month(dv_dateTime)]
dt[, dv_hour := lubridate::hour(dv_dateTime)]
dt[, dv_hms := hms::as_hms(dv_dateTime)]
dt <- gridCarbon::add_season(dt, 
                                 dateVar = "dv_dateTime",
                                 h = "S")
dt <- gridCarbon::add_peakPeriod(dt, 
                                      dateTime = "dv_dateTime")
  return(dt)
}

nzGenMix_wide_dt <- addDerivedVars(nzGenMix_wide_dt)
nzGenMix_dt <- addDerivedVars(orig_DT[!is.na(dv_dateTime)])

#message("Remove incomplete years to avoid weird things in plots.")

message("Filtered data range from: ", min(nzGenMix_wide_dt$dv_dateTime))
message("...to: ", max(nzGenMix_wide_dt$dv_dateTime))

rmdParams$plotCap <- paste0("Data: NZ EA Generation ",
                   min(nzGenMix_wide_dt$dv_date), " - ",
                   max(nzGenMix_wide_dt$dv_date),
                            "\nPlot: @dataknut \nCode: https://github.com/dataknut/gridCarbon")
```

>Note:
 >* the data covers more years than we need
 >* the data may contain partial years - BEWARE incomplete years in plots using annual totals or means across all months.

# Recreating DrSimEvans' plot

This looks like daily data.

 * Solar + wind (% of gen) vs coal + gas (& of gen))

```{r simEvansData}
# add together the % and totals we want (half-hourly)

nzGenMix_wide_dt[, dv_total := Coal + Diesel + Gas + Geo + Hydro + Solar + Wind + Wood]
nzGenMix_wide_dt[, dv_coal_gas := Coal + Gas]
nzGenMix_wide_dt[, dv_coal_gas_pc := 100 * dv_coal_gas/dv_total]
nzGenMix_wide_dt[, dv_solar_wind := Solar + Wind]
nzGenMix_wide_dt[, dv_solar_wind_pc := 100 * dv_solar_wind/dv_total]

# keep the vars we want for clarity
temp <- nzGenMix_wide_dt[, .(dv_dateTime, dv_coal_gas, dv_solar_wind,
                    dv_coal_gas_pc, dv_solar_wind_pc, dv_total)]

temp[, dv_date := lubridate::date(dv_dateTime)]

# aggregate to daily data for plotting
plotDT <- temp[,
               .(mean_dv_solar_wind_pc = mean(dv_solar_wind_pc),
                 mean_dv_coal_gas_pc = mean(dv_coal_gas_pc),
                 total_dv_coal_gas = sum(dv_coal_gas),
                 total_dv_solar_wind = sum(dv_solar_wind),
                 total_dv_total = sum(dv_total),
                 nObs = .N), # to check for days with < 48 half hours
               keyby = .(dv_date)
               ]
plotDT[, dv_year := lubridate::year(dv_date)] # for plots
plotDT[, total_dv_coal_gas_pc := 100 * total_dv_coal_gas/total_dv_total] # daily %
plotDT[, total_dv_solar_wind_pc := 100 * total_dv_solar_wind/total_dv_total]

message("Check for days with less than 48 hours - this will be truncated data due to DST breaks. We hate DST breaks")
table(plotDT$nObs)
```

Figure \@ref(fig:meanNZrenewablesVsfossilHalfHourPC) shows the mean half-hourly % generation by each type per day. This is slightly convoluted - it is the mean of the sum of the 48 daily half-hourly values. Unfold the code above for clarity.

The smoothed curves are estimated for each year. The lines terminate at the maximum value for the year. I'm still trying to decide if they tell us anything useful.

```{r meanNZrenewablesVsfossilHalfHourPC, fig.cap="Mean half-hourly % generation by each type per day"}

ggplot2::ggplot(plotDT[dv_year > 2011], aes(x = mean_dv_solar_wind_pc, 
                            y = mean_dv_coal_gas_pc,
                            colour = as.factor(dv_year),
                            alpha = dv_year)) +
  geom_point() +
  geom_smooth() +
  scale_colour_viridis_d(name = "Year") +
  guides(alpha = "none") +
  labs(x = "Solar & wind (mean % of half-hourly generation per day)",
       y = "Coal & gas (mean % of half-hourly generation per day)",
       caption = rmdParams$plotCap)

# save it
ggplot2::ggsave(filename = "meanNZrenewablesVsfossilHalfHourPC.png", 
                path = rmdParams$plotPath,
                height = 5)
```

Figure \@ref(fig:dailyNZpcGenMix) shows the percentage of daily generation by type. This is less convoluted as it is the sum of generation per day for the two categories (solar + wind vs gas + coal) as a % of total daily generation.

Again the smoothed curve is estimated for each year. 

```{r dailyNZpcGenMix, fig.cap="Percentage of daily generation by type"}
ggplot2::ggplot(plotDT[dv_year > 2011], aes(x = 100 * total_dv_solar_wind_pc, 
                            y = 100 * total_dv_coal_gas_pc,
                            colour = as.factor(dv_year),
                            alpha = dv_year)) +
  geom_point() +
  geom_smooth() +
  scale_colour_viridis_d(name = "Year") +
  guides(alpha = "none") +
  labs(x = "Solar & wind (% of total daily generation)",
       y = "Coal & gas (% of total daily generation)",
       caption = rmdParams$plotCap)

ggplot2::ggsave(filename = "dailyNZpcGenMix.png", 
                path = rmdParams$plotPath,
                height = 5)
```


## Half-hourly versions of the plot

Just cos we can... helpfully split into 'peak' and 'off peak' periods. 

Peak period definitions:

 * Morning 07:00 - 09:00
 * Daytime 09:00 - 16:00
 * Evening 16:00 - 21:00
 * Night - all other times

Again the smoothed curve is estimated for each year (and demand period). 

```{r halfHourlyPCgenByPeakPeriod, fig.cap="Percentage of half-hourly generation by type"}
ggplot2::ggplot(nzGenMix_wide_dt[dv_year > 2011], aes(x = dv_solar_wind_pc, 
                            y = dv_coal_gas_pc,
                            alpha = dv_year,
                            colour = as.factor(dv_year))) +
  geom_point() +
  facet_wrap(. ~ dv_peakPeriod) +
  geom_smooth() +
  scale_colour_viridis_d(name = "Year") +
  guides(alpha = "none") +
  labs(x = "Solar & wind (% of half-hourly generation)",
       y = "Coal & gas (% of half-hourly generation)",
       caption = rmdParams$plotCap)

ggplot2::ggsave(filename = "halfHourlyNZ_PCgenByPeakPeriod.png", 
                path = rmdParams$plotPath,
                height = 5)
```


# MWh generation trends

Values are in kWh per half hour https://www.emi.ea.govt.nz/Wholesale/Datasets/Generation/Generation_MD

```{r longTrendsDaily}

plotDT <- nzGenMix_dt[, 
                  .(dailyMWh = sum(as.numeric(kWh/1000))), # MWh -> MWh
                  keyby = .(dv_year, dv_date, dv_season, Fuel_Code)]

# annual sum for cross-check 
t <- plotDT[, .(sum_TWh = sum(dailyMWh/1000000)), keyby = .(Year = dv_year)]
t[, Year := as.factor(Year)]
gridCarbon::make_flexTable(t, caption = "Total TWh per year for cross-check")

# double check
```


Figure \@ref(fig:dailyGenBySource) shows individual trends.

```{r dailyGenBySource, fig.height = 6, fig.cap = "Trends in daily GWh generation"}
p <- ggplot2::ggplot(plotDT, aes(x = dv_date, y = dailyMWh/1000, colour = Fuel_Code)) +
  geom_line() +
  facet_grid(Fuel_Code ~ .) +
  scale_color_viridis_d(name = "Source") +
  labs(x = "Date",
       y = "GWh",
       caption = rmdParams$plotCap)
p


plotly::ggplotly(p)

ggplot2::ggsave(filename = "dailyNZGenBySource.png", 
                path = rmdParams$plotPath,
                height = 5)
```

Figure \@ref(fig:genTrendStackBySource) stacks them

```{r genTrendStackBySource, fig.cap = "Trends in daily GWh generation (stacked)"}
ggplot2::ggplot(plotDT, aes(x = dv_date, y = dailyMWh/1000, fill = Fuel_Code)) +
  geom_col(position = "stack") +
  scale_fill_viridis_d(name = "Source") +
  labs(x = "Date",
       y = "GWh",
       caption = rmdParams$plotCap)

ggplot2::ggsave(filename = "genNZTrendStackBySource.png", 
                path = rmdParams$plotPath
                )
```

## Renewable %

Defined how?

Renewable = wind + solar + hydro + geothermal + wood - but note that geothermal is not necessarily low carbon


```{r carbonStripesRenewable}

nzGenMix_wide_dt[, RENEWABLE := Geo + Hydro + Solar + Wind + Wood]

nzGenMix_wide_dt[, RENEWABLE_perc := 100*RENEWABLE/dv_total]

make_NgesoPlots <- function(dt, 
                            var = "RENEWABLE_perc", # defaults
                            lowColour = "green", 
                            highColour = "red", 
                            scaleLab = "Change me!", 
                            unit_lab = "??",
                            minMax = "max"){
  res <- list() # for the results
  res$tile <- ggplot2::ggplot(dt, aes(x = dv_date, 
                                      y = dv_hms, fill = get(var))) +
    geom_tile() +
    theme(legend.position = "bottom") +
    scale_x_date(date_labels="%b %Y",date_breaks  ="12 month") +
    theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
    scale_fill_continuous(name = paste0(scaleLab, " (", unit_lab, ")"),
                          high = highColour,
                          low = lowColour) +
    labs(x = "Date",
         y = "Half-hour",
         caption = rmdParams$plotCap)
  
  # if(minMax = "max"){
  #   res$tile <- res$tile + 
  #     scale_fill_continuous(name = paste0("Maximum ", scaleLab)
  #                           )
  # }
  #   if(minMax = "min"){
  #   res$tile <- res$tile + 
  #     scale_fill_continuous(name = paste0("Half-hourly ", scaleLab)
  #                           )
  #   }
  
  # line plot
  plotDT <- dt[,.(max = max(get(var)),
                  mean = mean(get(var)),
                  min = min(get(var))
  ),
  keyby = .(dv_date)]
  
  
  
  p <- ggplot2::ggplot(plotDT, aes(x = dv_date)) +
    theme(legend.position = "bottom") +
    scale_x_date(date_labels="%b %Y",date_breaks  ="12 month") +
    theme(axis.text.x = element_text(angle = 60, hjust = 1))
  
  if(minMax == "max"){
    yh <- max(plotDT$max)
    yLab <- paste0("Maximum half-hourly ", scaleLab, " (", unit_lab, ")")
    p <- p + geom_line(aes(y = max, colour = max)) +
      scale_color_continuous(name = yLab,
                             high = highColour,
                             low = lowColour) +
      geom_smooth(aes(y = max), colour = "grey") +
      geom_hline(yintercept = yh, colour = highColour) +
      annotate("text", x = mean(plotDT$dv_date), 
                      y = yh*1.05, 
               label = paste0("Maximum: ", round(yh), unit_lab))
    p <- p + labs(x = "Date",
                  y = yLab,
                  caption = rmdParams$plotCap
    )
  }
  
  if(minMax == "min"){
    yh <- min(plotDT$min)
    yLab <- paste0("Minimum half-hourly ", scaleLab, " (", unit_lab, ")")
    p <- p + geom_line(aes(y = min, colour = min)) +
      scale_color_continuous(name = yLab,
                             high = highColour,
                             low = lowColour) +
      geom_smooth(aes(y = min), colour = "grey") + 
  geom_hline(yintercept = yh, colour = lowColour) +
      annotate("text", x = mean(plotDT$dv_date), 
                      y = yh*1.05, 
               label = paste0("Minimum: ", round(yh), unit_lab))
    p <- p + labs(x = "Date",
                  y = yLab,
                  caption = rmdParams$plotCap
    )
  }
  res$line <- p
  
  return(res)
}

rpc <- make_NgesoPlots(nzGenMix_wide_dt, 
                       var = "RENEWABLE_perc",
                       lowColour = "red",
                       highColour = "green",
                       scaleLab = "Renewable generation",
                       unit = "%",
                       minMax = "max" # max line or min line?
)
rpc$tile
rpc$line
```

```{r carbonStripesRenewableRecent}

recentDateCut <- max(nzGenMix_wide_dt$dv_date - (6*30)) # roughyl 6 months

rpc <- make_NgesoPlots(nzGenMix_wide_dt[dv_date > recentDateCut], 
                       var = "RENEWABLE_perc",
                       lowColour = "red",
                       highColour = "green",
                       scaleLab = "Renewable generation",
                       minMax = "max",
                       unit_lab = "%"
)
rpc$tile +
  scale_x_date(date_labels="%b-%Y",date_breaks  ="1 month")

rpc$line +
  scale_x_date(date_labels="%b-%Y",date_breaks  ="1 month")

wpc <- nzGenMix_wide_dt[, .(mean_pc = mean(RENEWABLE_perc)), keyby = .(dv_hms, 
                                                                  Year = dv_year)]

ggplot2::ggplot(wpc, aes(x = dv_hms, alpha = Year, group = Year, y = mean_pc)) +
  geom_line() +
  labs(x = "Time of day",
       y = "% renewable generation")
```



## Wind %


```{r carbonStripesWind}
#plotDT <- gbGenMix_dt[, .()]
nzGenMix_wide_dt[, WIND_perc := 100*Wind/dv_total]
ci <- make_NgesoPlots(nzGenMix_wide_dt, 
                      var = "WIND_perc",
                      lowColour = "red",
                      highColour = "green",
                      scaleLab = "Wind",
                      minMax = "max",
                      unit_lab = "%"
)
ci$tile


ci$line

wpc <- nzGenMix_wide_dt[, .(mean_pc = mean(WIND_perc)), keyby = .(dv_hms, 
                                                                  Year = dv_year)]

ggplot2::ggplot(wpc, aes(x = dv_hms, alpha = Year, group = Year, y = mean_pc)) +
  geom_line() +
  labs(x = "Time of day",
       y = "% wind generation")
```

```{r carbonStripesWindRecent}
ci <- make_NgesoPlots(nzGenMix_wide_dt[dv_date > recentDateCut], 
                      var = "WIND_perc",
                      lowColour = "red",
                      highColour = "green",
                      scaleLab = "Wind",
                      minMax = "max",
                      unit_lab = "%"
)
ci$tile +
  scale_x_date(date_labels="%b-%Y",date_breaks  ="1 month")


ci$line +
  scale_x_date(date_labels="%b-%Y",date_breaks  ="1 month")
```

## Solar %


```{r carbonStripesSolar}
#plotDT <- gbGenMix_dt[, .()]
nzGenMix_wide_dt[, SOLAR_perc := 100*Solar/dv_total]
ci <- make_NgesoPlots(nzGenMix_wide_dt, 
                      var = "SOLAR_perc",
                      lowColour = "red",
                      highColour = "green",
                      scaleLab = "Solar",
                      minMax = "max",
                      unit_lab = "%"
)
ci$tile


ci$line

wpc <- nzGenMix_wide_dt[, .(mean_pc = mean(SOLAR_perc)), keyby = .(dv_hms, 
                                                                  Year = dv_year)]

ggplot2::ggplot(wpc, aes(x = dv_hms, alpha = Year, group = Year, y = mean_pc)) +
  geom_line() +
  labs(x = "Time of day",
       y = "% solar generation")
```

```{r carbonStripesSolarRecent}
ci <- make_NgesoPlots(nzGenMix_wide_dt[dv_date > recentDateCut], 
                      var = "SOLAR_perc",
                      lowColour = "red",
                      highColour = "green",
                      scaleLab = "Solar",
                      minMax = "max",
                      unit_lab = "%"
)
ci$tile +
  scale_x_date(date_labels="%b-%Y",date_breaks  ="1 month")


ci$line +
  scale_x_date(date_labels="%b-%Y",date_breaks  ="1 month")
```

## Hydro %

```{r carbonStripesHydro}
#plotDT <- gbGenMix_dt[, .()]
nzGenMix_wide_dt[, HYDRO_perc := 100*Hydro/dv_total]
ci <- make_NgesoPlots(nzGenMix_wide_dt, 
                      var = "HYDRO_perc",
                      lowColour = "red",
                      highColour = "green",
                      scaleLab = "Hydro",
                      minMax = "max",
                      unit_lab = "%"
)
ci$tile


ci$line

wpc <- nzGenMix_wide_dt[, .(mean_pc = mean(HYDRO_perc)), keyby = .(dv_hms, 
                                                                  Year = dv_year)]

ggplot2::ggplot(wpc, aes(x = dv_hms, alpha = Year, group = Year, y = mean_pc)) +
  geom_line() +
  labs(x = "Time of day",
       y = "% hydro generation")
```

```{r carbonStripesHydroRecent}
ci <- make_NgesoPlots(nzGenMix_wide_dt[dv_date > recentDateCut], 
                      var = "HYDRO_perc",
                      lowColour = "red",
                      highColour = "green",
                      scaleLab = "Hydro",
                      unit_lab = "%",
                      minMax = "max"
)
ci$tile +
  scale_x_date(date_labels="%b-%Y",date_breaks  ="1 month")


ci$line +
  scale_x_date(date_labels="%b-%Y",date_breaks  ="1 month")
```

## Geothermal %

```{r carbonStripesGeo}
#plotDT <- gbGenMix_dt[, .()]
nzGenMix_wide_dt[, GEO_perc := 100*Geo/dv_total]
ci <- make_NgesoPlots(nzGenMix_wide_dt, 
                      var = "GEO_perc",
                      lowColour = "red",
                      highColour = "green",
                      scaleLab = "Geothermal",
                      minMax = "max",
                      unit_lab = "%"
)
ci$tile


ci$line

wpc <- nzGenMix_wide_dt[, .(mean_pc = mean(GEO_perc)), keyby = .(dv_hms, 
                                                                  Year = dv_year)]

ggplot2::ggplot(wpc, aes(x = dv_hms, alpha = Year, group = Year, y = mean_pc)) +
  geom_line() +
  labs(x = "Time of day",
       y = "% geothermal generation")
```

```{r carbonStripesGeoRecent}
ci <- make_NgesoPlots(nzGenMix_wide_dt[dv_date > recentDateCut], 
                      var = "GEO_perc",
                      lowColour = "red",
                      highColour = "green",
                      scaleLab = "Geothermal",
                      minMax = "max",
                      unit_lab = "%"
)
ci$tile +
  scale_x_date(date_labels="%b-%Y",date_breaks  ="1 month")


ci$line +
  scale_x_date(date_labels="%b-%Y",date_breaks  ="1 month")
```


## Gas %

```{r carbonStripesGas}
#plotDT <- gbGenMix_dt[, .()]
nzGenMix_wide_dt[, GAS_perc := 100*Gas/dv_total]
ci <- make_NgesoPlots(nzGenMix_wide_dt, 
                      var = "GAS_perc",
                      lowColour = "green",
                      highColour = "red",
                      scaleLab = "Gas",
                      minMax = "max",
                      unit_lab = "%"
)
ci$tile


ci$line

wpc <- nzGenMix_wide_dt[, .(mean_pc = mean(GAS_perc)), keyby = .(dv_hms, 
                                                                  Year = dv_year)]

ggplot2::ggplot(wpc, aes(x = dv_hms, alpha = Year, group = Year, y = mean_pc)) +
  geom_line() +
  labs(x = "Time of day",
       y = "% gas generation")
```

```{r carbonStripesGasRecent}
ci <- make_NgesoPlots(nzGenMix_wide_dt[dv_date > recentDateCut], 
                      var = "GEO_perc",
                      lowColour = "green",
                      highColour = "red",
                      scaleLab = "Gas",
                      minMax = "max",
                      unit_lab = "%"
)
ci$tile +
  scale_x_date(date_labels="%b-%Y",date_breaks  ="1 month")


ci$line +
  scale_x_date(date_labels="%b-%Y",date_breaks  ="1 month")
```

## Coal %

```{r carbonStripesCoal}
#plotDT <- gbGenMix_dt[, .()]
nzGenMix_wide_dt[, COAL_perc := 100*Coal/dv_total]
ci <- make_NgesoPlots(nzGenMix_wide_dt, 
                      var = "COAL_perc",
                      lowColour = "green",
                      highColour = "red",
                      scaleLab = "Coal",
                      minMax = "max",
                      unit_lab = "%"
)
ci$tile


ci$line

wpc <- nzGenMix_wide_dt[, .(mean_pc = mean(COAL_perc)), keyby = .(dv_hms, 
                                                                  Year = dv_year)]

ggplot2::ggplot(wpc, aes(x = dv_hms, alpha = Year, group = Year, y = mean_pc)) +
  geom_line() +
  labs(x = "Time of day",
       y = "% coal generation")
```

```{r carbonStripesCoalRecent}
ci <- make_NgesoPlots(nzGenMix_wide_dt[dv_date > recentDateCut], 
                      var = "COAL_perc",
                      lowColour = "green",
                      highColour = "red",
                      scaleLab = "Coal",
                      minMax = "max",
                      unit_lab = "%"
)
ci$tile +
  scale_x_date(date_labels="%b-%Y",date_breaks  ="1 month")


ci$line +
  scale_x_date(date_labels="%b-%Y",date_breaks  ="1 month")
```


# Carbon intensity trends

> XX need to create a half-hourly `CARBON_INTENSITY` variable XX
> Requires:
> * per-fuel CI

Currently: https://environment.govt.nz/assets/publications/Measuring-emissions-guidance-August-2022/Emission-factors-workbook-Measuring-emissions-guidance-August-2022.xlsx 

"The grid-average emission factor best reflects the carbon dioxide equivalent emissions
associated with the generation of a unit of electricity purchased from the national grid in New Zealand in 2020"

"We calculate purchased electricity emission factors on a calendar-year basis and based on the average grid mix of generation types for calendar years. The emission factor accounts for the emissions from fuel combustion at thermal power stations and fugitive emissions from the generation of geothermal electricity. Thermal electricity is generated by burning fossil fuels.

The emission factor for purchased grid-average electricity does not include transmission
and distribution losses. "

"This emission factor also doesn’t reflect the real-world factors that influence the carbon intensity of the grid such as time of year, time of day and geographical area. Therefore, a grid -average emission factor may over or underestimate your organisation’s GHG emissions."

2020 value = 0.120 kg CO2/kWh

Instead use a [model derived from the GB MWh generation mix for 2023](https://dataknut.github.io/gridCarbon/gbGenMixTrends.html#tab:plotEstimates).

Coefficients:

 * Intercept: 41.231171 
 * GAS_perc: 3.571570 
 * COAL_perc: 10.050410
 * WIND_perc: -0.451285
 * HYDRO_perc:  1.147793 # why is hydro +ve?
 * SOLAR_perc: -0.335122
 
This is (obviously) completely broken as HYDRO_perc has a +ve coefficient, the model does not include a coefficient for Geothermal or Wood and the GB mix includes Nuclear, Interconnection and Storage. #YMMV

> Need to fix :-)

But let's imagine it is valid... what do we see?

```{r estimateCarbonIntensity}
# should be annual and fuel specific
# 2020 value

nzGenMix_wide_dt[, COAL_perc := Coal/dv_total]
nzGenMix_wide_dt[, GAS_perc := Gas/dv_total]
nzGenMix_wide_dt[, COAL_perc := Coal/dv_total]

nzGenMix_wide_dt[, CARBON_INTENSITY := 41.231171 +
                   GAS_perc   *  3.571570   + 
                   COAL_perc   * 10.050410   +
                   WIND_perc  * -0.451285  +
                   HYDRO_perc *  1.147793 + # why is hydro +ve?
                   SOLAR_perc * -0.335122]

summary(nzGenMix_wide_dt$CARBON_INTENSITY)
```


## Overall trends

Seasons:

 * Summer: Jun - Aug
 * Autumn: Sept - Nov
 * Winter: Dec - Feb
 * Spring: Mar - May

```{r meanCiTrendBySeason, fig.cap="Mean half-hourly carbon intensity per day"}

plotDT <- nzGenMix_wide_dt[, .(mean_CI = mean(CARBON_INTENSITY)), keyby = .(dv_year, dv_date, dv_season)]

ggplot2::ggplot(plotDT[dv_year > 2011], aes(x = dv_date, y = mean_CI, colour = dv_season)) +
  geom_point() +
  scale_color_viridis_d(name = "Season") +
  theme(legend.position = "bottom") +
  labs(x = "Date",
       y = "Mean g CO2e/kWh",
       caption = rmdParams$plotCap)

ggplot2::ggsave(filename = "meanCiTrendBySeason.png", 
                path = rmdParams$plotPath
                )
```
Re-draw as a boxplot of mean daily CI by month - plotted at month start. \@ref(fig:CIboxplot) shows outliers nicely.

```{r CIboxplot, fig.cap="Mean daily CI boxplots per month"}
plotDT <- nzGenMix_wide_dt[, .(mean_CI = mean(CARBON_INTENSITY)), 
                     keyby = .(dv_year, 
                               dv_date,
                               dv_season)]
plotDT[, dv_month := lubridate::floor_date(dv_date, unit = "months"),]
ggplot2::ggplot(plotDT[dv_year > 2011], aes(x = dv_month, 
                                            y = mean_CI, 
                                            group = dv_month,
                                            colour = dv_season)) +
  geom_boxplot() +
  scale_color_viridis_d(name = "Season") +
  theme(legend.position = "bottom") +
  labs(x = "Month",
       y = "Mean g CO2e/kWh",
       caption = rmdParams$plotCap)

ggplot2::ggsave(filename = "meanCiTrendByMonthSeason.png", 
                path = rmdParams$plotPath
                )
```

Figure \@ref(fig:carbonStripesCI) shows half-hourly carbon intensity over time.

```{r carbonStripesCI, fig.cap="Half-hourly carbon intensity over time"}
#plotDT <- nzGenMix_wide_dt[, .()]
ci <- make_NgesoPlots(nzGenMix_wide_dt, 
                      var = "CARBON_INTENSITY",
                      lowColour = "green",
                      highColour = "brown",
                      scaleLab = "Carbon intensity (g CO2/MW)",
                      minMax = "min"
)
ci$tile
```


```{r carbonStripesCILine}
ci$line
```

```{r carbonStripesCIRecent}
ci <- make_NgesoPlots(nzGenMix_wide_dt[dv_year > 2021], 
                      var = "CARBON_INTENSITY",
                      lowColour = "green",
                      highColour = "brown",
                      scaleLab = "Carbon intensity (g CO2/MW)",
                      minMax = "min"
)
ci$tile +
  scale_x_date(date_labels="%b-%Y",date_breaks  ="3 month")


ci$line +
  scale_x_date(date_labels="%b-%Y",date_breaks  ="3 month")
```


## Annual by peak period

Mean carbon intensity per year and season within peak period.

Figure \@ref(fig:annualmeanCIByPeak) suggests evening peak periods still have slightly higher carbon intensity and the shape of the reduction curves differ by season although rather less by period. Interestingly the sustained reduction in carbon intensity in Summer has leveled off.

```{r annualmeanCIByPeak, fig.cap="Mean half-hourly carbon intensity by peak period and season 2012-2022"}
plotDT <- nzGenMix_wide_dt[, .(mean_CI = mean(CARBON_INTENSITY)), keyby = .(dv_year, dv_peakPeriod, dv_season)]

ggplot2::ggplot(plotDT, aes(x = dv_year, y = mean_CI, colour = dv_peakPeriod)) +
  geom_line() +
  scale_color_viridis_d(name = "Peak period") +
  facet_wrap(. ~ dv_season) +
  labs(x = "Year",
       y = "Mean g CO2e/kWh",
       caption = rmdParams$plotCap)

ggplot2::ggsave(filename = "annualMeanCIByPeak.png", 
                path = rmdParams$plotPath
                )
```

# Renewables and overall generation

Do we see a relationship between renewables generating and peak demand? This will be mediated by the way the electricity market works. 

We may find wind curtailment (not visible here) at low demand periods where nuclear can't be shut off.

First, what is the general average shape of carbon intensity and renewable generation?

Figure \@ref(fig:ciProfile) shows that although the mean half-hourly carbon intensity had fallen over time (with the effect of solar in summer particularly noticeable), the morning and evening peaks are still relatively more carbon intense as demand overtakes the available renewable supply.

```{r ciProfile, fig.cap = "Mean half-hourly carbon intensity by year and season"}
plotDT <- nzGenMix_wide_dt[dv_year > 2011, .(mean_ci = mean(CARBON_INTENSITY),
                          mean_renewables_MW = mean(RENEWABLE),
                          mean_renewables_pc = mean(RENEWABLE_perc)),
                          keyby = .(dv_year, dv_hms, dv_season)]

ggplot2::ggplot(plotDT, aes(x = dv_hms, y = mean_ci, 
                            alpha = dv_year,
                            colour = dv_year,
                            group = dv_year)) +
  geom_line() +
  scale_alpha_continuous(name="Year") +
  scale_color_continuous(name = "Year",
                         low = "grey", 
                         high = "#3CBAC6") + # UoS from Marine palette
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  facet_wrap(. ~ dv_season) +
  labs(x = "Time of day",
       y = "Mean half-hourly carbon intensity",
       caption = rmdParams$plotCap)
```

```{r meanrenewables, fig.cap="Mean half-hourly renewable generation by year and season"}
ggplot2::ggplot(plotDT, aes(x = dv_hms, y = mean_renewables_MW/1000, 
                            colour = dv_year, 
                            alpha = dv_year,
                            group = dv_year)) +
  geom_line() +
  scale_alpha_continuous(name = "Year") +
  scale_color_continuous(name = "Year",
                         low = "grey", 
                         high = "#3CBAC6") + # UoS from Marine palette
  facet_grid(dv_season ~ .) +
    labs(x = "Time of day",
       y = "Mean renewables (GW)",
       caption = rmdParams$plotCap)

ggplot2::ggplot(plotDT, aes(x = dv_hms, y = mean_renewables_pc, 
                            colour = dv_year, 
                            alpha = dv_year,
                            group = dv_year)) +
  geom_line() +
  scale_alpha_continuous(name = "Year") +
  scale_color_continuous(name = "Year",
                         low = "grey", 
                         high = "#3CBAC6") + # UoS from Marine palette
  facet_grid(dv_season ~ .) +
    labs(x = "Time of day",
       y = "Mean % renewables (%)",
       caption = rmdParams$plotCap)
```


```{r renewableVsGeneration}
ggplot2::ggplot(nzGenMix_wide_dt, aes(x = RENEWABLE, y = dv_total)) +
  geom_point() +
  facet_wrap(. ~ dv_year) +
  theme(axis.text.x = element_text(angle = 45, hjust=1))
```

\@ref(fig:meanRenewablesByToD) shows the rise of both wind and solar (mid-day peak)...

> Beware incomplete yearss

```{r meanRenewablesByToD, fig.cap="Trends in mean half-hourly renewable generation by time of day and season"}
nzGenMix_wide_dt[, GENERATION := dv_total]
plotDT <- nzGenMix_wide_dt[, .(mean_renewables = mean(RENEWABLE),
                          mean_solar = mean(Solar),
                 mean_generation = mean(GENERATION)),
             keyby = .(dv_year, dv_hms, dv_peakPeriod, dv_season)]

ggplot2::ggplot(plotDT[dv_year > 2016], aes(x = dv_hms, 
                                         colour = dv_peakPeriod,
                                         alpha = dv_year,
                                         group = dv_year)) +
  geom_line(aes(y = mean_renewables/1000)) +
  scale_alpha_continuous(name = "Year") +
  scale_color_discrete(name = "Peak period") +
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  facet_wrap(. ~ dv_season) +
  labs(x = "Time of day",
       y = "Mean renewables (GW)",
       caption = rmdParams$plotCap)
```

What would have happened if we increased solar generation in 2022 by a factor of 10? We would have needed some storage in Spring & Summer...

```{r meanGenByToD, fig.cap="Comparing mean total generation & 10 * solar generation by time of day for 2020"}


ggplot2::ggplot(plotDT[dv_year == 2022], aes(x = dv_hms,
                                             colour = dv_peakPeriod,
                                             group = dv_peakPeriod)) +
  geom_point(aes(y = mean_generation/1000)) +
  geom_line(aes(y = mean_solar/1000 * 10), linetype = "dotdash") +
  scale_color_discrete(name = "Peak period") +
    theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  facet_wrap(. ~ dv_season) +
  labs(x = "Time of day",
       y = "Mean generation (GW)",
       caption = rmdParams$plotCap)
```

YMMV on \@ref(fig:meanRenewablesVsMeanGenSeason)

```{r meanRenewablesVsMeanGenSeason, fig.cap="Mean renewables vs Mean total generation"}
plotDT <- nzGenMix_wide_dt[, .(mean_renewables = mean(RENEWABLE),
                 mean_generation = mean(GENERATION)),
             keyby = .(dv_year, dv_hms, dv_peakPeriod, dv_season)]

ggplot2::ggplot(plotDT[dv_year > 2016], aes(x = mean_generation/1000 , y = mean_renewables/1000,
                                         colour = dv_peakPeriod)) +
  geom_point() +
  scale_color_discrete(name = "Period") +
  facet_grid(dv_season ~ dv_year) +
  labs(x = "Mean total generation (GW)",
       y = "Mean renewables (GW)",
       caption = rmdParams$plotCap)
```

# Summary

That's it. 

You might want to look at recent academic research on this topic:

 * [@staffell_measuring_2017]
 * [@staffell_increasing_2018]

# Annex

## Data descriptors

### NZ gen mix data

```{r skimNZGen}
skimr::skim(nzGenMix_wide_dt)
```

## R environment

Packages etc:

 * base R [@baseR]
 * bookdown [@bookdown]
 * data.table [@data.table]
 * ggplot2 [@ggplot2]
 * here [@here]
 * hms [@hms]
 * knitr [@knitr]
 * lubridate [@lubridate]
 * rmarkdown [@rmarkdown]

# References
